Opening up the court (surface) in tennis grand slams

### 2018-09-16 contest/surface-type/paper-tennis.Rmd
kayla edits

2. Examine fine grained attributes such as winners, aces, and unforced errors across the four surfaces

@@

There has been little work on quantifying the high-level effect of court surface in tennis issue (1). Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001]. @knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass compared to other courts. Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].

We examine the match attributes more unique to tennis to address issue (2). These attributes include winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).  In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].

Finally, to assess how individual players differ in performance across surfaces, issue (3), we take two approaches.  The first is to use a hierarchical model with fixed and random effects, taking into account player and tournament attributes using the complete data set.  This allows us to leverage tournament information to model players with few matches played and to help account for noise.  Our second approach is to model performance for specific players individually. Since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with tennis-specific feature attributes to examine the effects individually.  In both cases, we examine whether our models pass "common sense" tests like how the models in @thomas2013 show that commonly well known hockey players also have high status in the model.  We also examine whether these players do have court surface effects. For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.  However, when the data becomes available, we would like to test these models on the 2018 grand slam results. For model selection, we use AIC to compare models to one another [@wasserman2004].

@@

```{r table, fig.pos = 'H'}
data(gs_partial_players)

top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
  summarize(matches_played = n())
matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
  summarize(matches_played = sum(matches_played))
tab <- spread(matches_played_slam, key = name, value = matches_played)
total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
colnames(total_df) <- colnames(tab)
tab <- rbind(tab, total_df)

@@

```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center', fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}

@@

```{r, out.width="70%", fig.width=7, fig.height=3.5, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter() +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament", values = tournament_colors)

@@

We considered a variety of models incorporating effects as either fixed or random. That is $\pi_i = \beta X_i + \delta_i$, where $\pi_i = P(Y_i = 1)$ (i.e. a player win), $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years. Based on AIC, we found the model that included a player-level term for each tournament, but excluded fixed terms for country, was the best fit for the data. The models are summarized in Table \ref{tab:model.sums}.

@@

```{r, fig.height = 3.5, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "none") + 
  xlab("") +
  ylab("Estimate") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

random.effects.plot = ran_tourn %>%
  gather(., tourn, estimate, -model) %>%
  mutate(., tournament = gsub("tournament", "", tourn)) %>%
  ggplot(., aes(x = tournament, y = estimate, fill = model)) +
  geom_violin() + 
  my_theme +
  scale_fill_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Random effects") +
  theme(legend.position = "bottom") +
  guides(col = guide_legend(nrow=3))

@@

grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5),
             top = textGrob("`ind_no_ioc' model",
                            gp=gpar(fontsize=20, fontfamily="serif")))
```

@@

```{r, fig.height = 3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Aces") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  coord_flip() +
  theme(legend.position = "none")

@@

At the highest level, we find that the general point distribution seems to be the same for all court surfaces. There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g. that Spaniards perform better on clay). We do not find evidence for a home court advantage across all players in the grand slams. We also find that the FO seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.

Using a hierarchical model approach, we do not see strong individual effects for Williams, Federer, and Nadal in predicting wins. We do, however, see differences for these players in hierarchical models to predict aces, net winners, and UEs, and that these differences are largely consistent with common tennis knowledge.

>>>>

2. Examine tennis specific attributes such as winners, aces, and unforced errors across the four surfaces

@@

There has been little work on quantifying the high-level effect of court surface in tennis (issue 1). Existing literature provides methods for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001]. @knottenbelt2012 take court surface into account in their model, but do not compare the results of one surface to another, and @pollard2007 conclude that players have more injuries on grass compared to other courts. Results from studies performed in other sports show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010].

We examine the match attributes more unique to tennis to address (issue 2). These attributes include winners (when a player hits a shot that the opponent is unable to return), aces (a serve an opponent is unable to return), and unforced errors (UE) (when a player makes a simple mistake resulting in a loss of the point).  In general, players hope to maximize the winner to UE ratio (W/UE), which may lead to playing more aggressively on faster courts and more conservatively on slower courts such as clay [@theage2007].

Finally, to assess how individual players differ in performance across surfaces, (issue 3), we take two approaches.  The first is to use a hierarchical model with fixed and random effects, taking into account player and tournament attributes using the complete data set.  This allows us to leverage tournament information to model players with few matches played and to help account for noise.  Our second approach is to model performance for specific players individually. Since Nadal, Federer, and Williams are extremely popular and have had much success in the 2013-2017 time range, we have enough data points with tennis-specific feature attributes to examine the effects individually.  In both cases, we examine whether our models pass "common sense" tests like how the models in @thomas2013 show that commonly well known hockey players also have high status in the model.  We also examine whether these players do have court surface effects. For the individual models, we do not split the data into training and test sets due to the relatively low number of observations for the individuals.  However, when the data becomes available, we would like to test these models on the 2018 grand slam results. For model selection, we use AIC to compare models to one another [@wasserman2004].

@@

```{r table, fig.pos = 'H'}
top_three <- dplyr::filter(gs_players, name %in% c("Rafael Nadal", "Roger Federer", "Serena Williams"))
matches_played_slam_year <- top_three %>% group_by(name, tournament, year) %>%
  summarize(matches_played = n())
matches_played_slam <- matches_played_slam_year %>% group_by(name, tournament) %>%
  summarize(matches_played = sum(matches_played))
tab <- spread(matches_played_slam, key = name, value = matches_played)
total_df <- data.frame(tournament = "Total", as.list(colSums(tab[,-1])))
colnames(total_df) <- colnames(tab)
tab <- rbind(tab, total_df)

@@

```{r, fig.height=3.5, fig.width=10, fig.align='center', fig.pos = 'H', fig.cap = "\\label{fig:home-court} A home court advantage does not appear to exist in any of the grand slams after we account for the number of players from a given country."}
france_wins <- filter(gs_players, ioc == "FRA", round < "QF") %>% 
  group_by(tournament, round) %>% 
  summarise(win_prop = mean(did_win),
            win_prop_se = sqrt(win_prop*(1-win_prop)/n()))

@@

```{r, fig.height=3.8, fig.width=8, out.width="80%", fig.align='center',  fig.pos = 'H',  fig.cap = "\\label{fig:spain-clay} We find that spaniards win more often at the French Open than other slams, even though their ranking is, on average, lower in the FO."}

@@

```{r, out.width="70%", fig.width=7, fig.height=3.4, fig.align='center', fig.cap= "\\label{fig:ue-ace}We see Wim. matches tend to have more aces and fewer unforced errors than the other tournaments."}
ggplot(gs_partial, aes(w_n_ue + w_n_ue, w_n_aces + w_n_aces,
                           label = tournament, color=tournament)) + 
  geom_jitter() +
  labs( title = "Distribution of errors and aces differ by tournament",
        x = "Total unforced errors", y = "Total aces") +
  my_theme + scale_color_manual("Tournament", values = tournament_colors)

@@

We considered a variety of models incorporating effects as either fixed or random. That is $E[Y_{\% pts. won}|X] = \beta X$, where $E[\pi_i|X_i, \delta_i] = \beta X_i + \delta_i$ (i.e. a player win), $X_i$ are the fixed effects included in the model, and $\delta_i$ is a random effect across either players or years. Based on AIC, we found the model that included a player-level term for each tournament, but excluded fixed terms for country, was the best fit for the data. The models are summarized in Table \ref{tab:model.sums}.

@@

```{r, fig.height = 3.8, fig.width=10, fig.cap ="\\label{fig:mixed-mods-ests}Estimated fixed effects for all seven models (left plot) and player-level effects for Williams, Federer and Nadal under the ind_no_ioc model."}
fixed.effects.plot = model.ests %>%
  filter(., !grepl("ioc", term)) %>%
  ggplot(., aes(x = term, y = estimate, col = model, ymin = estimate - 2*std.error, ymax = estimate + 2*std.error)) +
  geom_errorbar(width = .7, position = position_dodge()) +
  my_theme +
  # coord_flip() +
  theme(axis.text.x=element_text(angle = 30, hjust = 1)) +
  ggtitle("Fixed effects") +
  scale_color_brewer("",palette = "Dark2") +
  theme(legend.position = "bottom", legend.text=element_text(size=8),
        # legend.margin=margin(0, 0, 0, 0),
        legend.margin=margin(t = 0, unit='cm'),
        legend.box.margin=margin(-10,-10,-10,-10)) + 
  xlab("") +
  ylab("Estimate") +
  scale_x_discrete("", breaks = waiver(), labels = c("Intercept", "ATP", "Late round", "Op Rank (log)", "Rank (log)", "French", "US", "Wimbledon", "2014", "2015", "2016", "2017"))+
  guides(col = guide_legend(nrow=2))

@@

random.effects.plot = ran_tourn %>%
  gather(., tourn, estimate, -model) %>%
  mutate(., tournament = gsub("tournament", "", tourn)) %>%
  ggplot(., aes(x = tournament, y = estimate, fill = model)) +
  geom_violin() + 
  my_theme +
  scale_fill_brewer(palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Random effects") +
  theme(legend.position = "bottom", legend.text=element_text(size=8)) +
  guides(col = guide_legend(nrow=3))

@@

grid.arrange(fixed.effects.plot, top_three_ranef, nrow = 1, widths = c(.5, .5),
             top = textGrob("Parameter estimates from logistic model",
                            gp=gpar(fontsize=20, fontfamily="serif")))
```

@@

```{r, fig.height = 3.3, fig.width=10, fig.cap = "\\label{fig:mixed-mods-stats}Estimated player-level effects for Williams, Federer and Nadal when modeling aces, net winners, and unforced errors."}
aces_ranef = as.data.frame(ranef(n_aces_mod, condVar = TRUE)) %>%
  mutate(tournament = gsub("tournament", "", term)) %>%
  filter(., grp == "Serena Williams" | grp == "Rafael Nadal" | grp == "Roger Federer") %>%
  ggplot(., aes(x = as.character(grp), y = condval, ymin = condval- 2*condsd, ymax = condval + 2*condsd, col = tournament)) +
  geom_errorbar(height = .5, position = position_dodge()) +
  ylab("Estimate") + 
  xlab("") +
  ggtitle("Aces") + 
  scale_color_manual("", values=tournament_colors) +
  my_theme +
  coord_flip() +
  theme(legend.position = "none")

@@

At the highest level, we find that the general point distribution seems to be the same for all court surfaces. There may be some differences in player performance on the different court surfaces, but these differences are not definitive (e.g. that Spaniards perform better on clay). We do not find evidence for a home court advantage across all players in the grand slams. We also find that the French Open seems to be the most different out of the four grand slams, but Wimbledon also displays some unique characteristics with regards to low amounts of UEs.

Using a hierarchical model approach, we do not see strong individual effects for Williams, Federer, and Nadal in predicting wins. We do, however, see differences for these players in hierarchical models to predict aces, net winners, and unforced errors, and that these differences are largely consistent with common tennis knowledge.
